SSA based global code motion of identical computations

### 2017-04-15 acmart-pacmpl-template.tex
move example in introduction

This paper describes code-motion of identical computations, a technique that
uses the information computed for \PRE{} to detect identical computations but
has a transformation phase whose goal differs from \PRE{}: it removes identical
computations from different branches of execution.  These identical computations
in different branches of execution are not redundant computations at run-time
and the number of run-time computations is not reduced. It is not a redundancy
elimination pass, and thus it has different cost function and heuristics than
\PRE{} or CSE. It is more similar to global code scheduling
\cite{dragonbook,click1995global} in the sense that it will only move
computations. The goals of this code motion are:

\begin{itemize}
\item to reduce the code size of the program;
\item to improve function inlining heuristics: functions become cheaper to
  inline when their code size decreases because inliner heuristics in \LLVM{}
  depends on instruction count.
\item to expose more instruction level parallelism: by hoisting identical
  computations to be executed earlier, instruction schedulers can move heavy
  computations earlier in order to avoid pipeline bubbles;
\item to help out-of-order processors with speculative execution of branches: by
  hoisting/sinking expressions out of branches, code hoisting can effectively
  reduce the amount of code to be speculatively executed and hence reduce the
  critical path;
\item to reduce interference/register pressure;
\item for SIMD architectures, which execute all branches, it will reduce the
  total number of instructions to be executed.
\item to improve passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted or sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of code
hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
hoisting from identical basic blocks: the instructions of two sibling basic
blocks are read in the same time, and all the instructions of the blocks are
hoisted to the common parent block as long as the compiler is able to prove that
the instructions are equivalent.  This implementation does not allow for an easy
extension: first in terms of compilation time overhead the implementation would
become quadratic in number of instructions to bisimulate, and second the
equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
would need to be rewritten to be more general, leading to using a mechanism
similar to the idea described in this paper based on \GVN{}.

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
are computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

Rosen \cite{rosen1988global} also introduced moving computations from successors
using value numbering. Their algorithm iterates on computations of same rank and
move the code with identical computations from the sibling branch to a common
dominator if they are very busy \cite{steven1997advanced}.

@@

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

>>>>

This paper describes code-motion of identical computations, a technique that
uses the information computed for \PRE{} to detect identical computations but
has a transformation phase whose goal differs from \PRE{}: it removes identical
computations from different branches of execution.  These identical computations
in different branches of execution are not redundant computations at run-time
and the number of run-time computations is not reduced. It is not a redundancy
elimination pass, and thus it has different cost function and heuristics than
\PRE{} or CSE. It is more similar to global code scheduling
\cite{dragonbook,click1995global} in the sense that it will only move
computations. We will now consider an example that explains how code hoisting can
also improve performance.

In fact there are several advantages of code-motion of this kind:
\begin{itemize}
\item It reduces the code size of the program;
\item It improves function inlining heuristics: functions become cheaper to
  inline when their code size decreases because inliner heuristics in \LLVM{}
  depends on instruction count.
\item It expose more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item It reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting/sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item It reduces interference/register pressure when appropriate cost-model is
  applied. We have used a cost model in our implementation which hoists/sinks only
  when register pressure would potentially reduce.
\item For SIMD architectures, which execute all branches, it will reduce the
  total number of instructions to be executed.
\item It improves passes that do not work well with branches:
  \begin{itemize}
  \item to improve loop vectorization by reducing a loop with control flow to a
    loop with a single BB, should all the instructions in a conditional get
    hoisted or sinked;
  \item to enable more loop invariant code motion (LICM): as LICM does not
    reason about instructions in the context of loops with conditional branches,
    code-hoisting is needed to move instructions out of conditional expressions
    and expose them to LICM.
  \end{itemize}
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic blocks: the instructions of two sibling basic blocks are read in the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

Rosen \cite{rosen1988global} also introduced moving computations from successors
but using value numbering. Their algorithm iterates on computations of same rank
and move the code with identical computations from the sibling branch to a
common dominator if they are very busy \cite{steven1997advanced}. We also
incorporate ranking of expressions in our implementation.

@@

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is able to prove are dependent.
\end{itemize}
